A Large and Variable Leading Tail of Helium in a Sub-Saturn Undergoing Runaway Inflation

### 2023-07-13 paper/paper1/ms.tex
Grammarly the whole thing, closes #39

\begin{abstract}
    Atmospheric escape shapes the fate of exoplanets, with statistical evidence for transformative mass loss imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10833 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of up to 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67~b. The 13.8 hours of on-sky integration time over 39 nights sample the entire planet orbit, uncovering excess Helium absorption preceding the transit by up to 130 planetary radii in a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe, and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary winds.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We contrast two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

@@

Atmospheric escape appears to be a normal phase for exoplanets of a certain size and insolation, statistically imprinted in the dearth of 1.5$-$2.0~$R_\oplus$ planets within 100 day periods \citep{2017AJ....154..109F}.  The leading candidate physical mechanisms for this rapid transition between mini-Neptunes and super-Earths include photoevaporation \citep{2013ApJ...775..105O,2013ApJ...776....2L,2017ApJ...847...29O} and core-powered mass loss \citep{2019MNRAS.487...24G,2023arXiv230200009B}.  Whatever the cause, some large fraction of planets undergo transformative atmospheric escape and the signal should be widely discernable.  Such signals have been searched for, and increasingly found, in many transiting planet systems with at least 28 detections to date \citep{2022arXiv221116243D}.

@@

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may either form \emph{in-situ} or effectively migrate to close-in orbital separations \citep{2018ARA&A..56..175D}, but once they arrive, the intense irradiation overheats the planet.  This heating leads to runaway inflation and ultimately complete disintegration.  In this scenario, the inflationary half-life becomes so short that the probability of observing members in the class decreases sharply with density, causing the apparent lack of inflated sub-Saturns \citep{2023ApJ...945L..36T}.

@@

Recently, \citet{2023ApJ...945L..36T} showed that hot Saturns can undergo catastrophic erosion by stellar X-ray and Extreme UV (XUV) photoevaporative mass loss. Planets with densities less than $\sim$0.1 g~cm$^{-3}$ achieve mass loss rates of up to $10^3$ $M_\oplus /$Gyr setting up a positive feedback loop: the planet increases in radius, overflowing its Roche lobe, fueling greater mass loss, and increasing in radius even further.  This vicious cycle systematically depopulates the mass-radius plane with a cliff defined by the $\rho_p \sim$0.1 g~cm$^{-3}$ dividing line.

@@

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2000ApJ...537..916S,2003Natur.422..143V,2012ApJ...751...86J,2018NatAs...2..714Y,2018ApJ...855L..11O,2018Natur.557...68S, 2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections \citep{2022arXiv221116243D}\footnote{See also Table~S1 of \citealt{doi:10.1126/sciadv.adf8736} for a compilation of exoplanets with detections and non-detections of the helium excess.}.  The sample of detections includes both hot Jupiters and lower mass planets, but lacks inflated hot Saturns owing to their intrinsic rarity below the 0.1 g~cm$^{-3}$ threshold. An observational picture of mass loss in inflated hot Saturns appears to be lacking for this reason.

@@

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis}, we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in Section \ref{secHeAnalysis}, with analysis of the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with 1D Parker wind models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

The Habitable Zone Planet Finder Spectrograph \citep[HPF;][]{2012SPIE.8446E..1SM,2014SPIE.9147E..1GM, 2019Optic...6..233M} on the queue-scheduled 10-meter \emph{Hobby-Eberly Telescope} \citep[HET;][]{1998SPIE.3352...34R} operates in the near-IR from $8100-12800~$\AA~ spanning the \textit{z}, \textit{Y}, and \textit{J} bands at spectral resolving power $R=55,000$. The HET fixed-elevation design \citep{2007PASP..119..556S} limits the observability of \object{HAT-P-67} to less than 1 hour ``tracks'' for a fixed range of hour angles before (east track) and after (west track) the star transits the meridian.  Whereas conventional steerable telescopes could conduct continuous point-and-stare observations of \object{HAT-P-67} for hours, HET cannot.  In practice, this limitation means that in-transit and out-of-transit observational phases were rarely possible on the same night.  Instead, we organized the observations into four campaigns to coincide with \object{HAT-P-67 b} transits on UT dates 2020 April 28, 2020; May 22, 2020; June 15, 2020; and 2022 April 29.  These campaigns have out-of-transit observations at least one night before and one night after, and often two nights on either side of the transit.  Two more transit snapshots were obtained in 2022 June-July without the accompanying visits immediately before and after.  The in-transit campaigns had up to 14 exposures per HET track, with integration times between 5-8.5 minutes.  We also obtained random-in-phase lower-priority ``P1-P4'' reconnaissance observations \citep{2007PASP..119..556S}.  These out-of-transit snapshot observations typically received 4 or fewer individual exposures.  We observed HAT-P-67 with HPF for a total of 41 visits on 39 unique nights, with two of those nights observing both the east and west HET tracks.  The total on-source integration time exceeds 13.8 hours, with $N=152$ individual spectra possessing typical signal-to-noise ratio of 80 per pixel.  Table \ref{tabHPFLog} summarizes the log of observations.

The observation schedule was strategically restricted to the spring season when the Barycentric Earth Radial Velocity (BERV) would Doppler shift telluric absorption lines sufficiently far away from the core of the Helium 10833 \AA~ feature \citep{2022ApJ...939L..11S}.  The small BERV still means the redward Helium line wing has significant telluric contamination between 10834$-$10836 \AA, but the core and blue line wings appear relatively pristine.  Sky emission lines land in this region, but are more easily mitigated by HPF's simultaneous sky reference fiber.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}, well-separated from the planet-host star and not a source of contamination for the HPF observations.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ mag source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation. TESS Sectors 51-53 do not show as conspicuous a modulation signal but still show some ostensibly stellar variability.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone the periods hover around 5.9 days.  The signal is weak enough, especially in Sectors 51-53, that \emph{TESS} instrumental systematics may contribute some of the out-of-transit modulation that we see in Figure \ref{fig:TESSoverview}.

We can independently constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i_\star}$, and the observation of the planet's orbital inclination $i_p\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We assume spin-orbit alignment, $i_\star \sim i_p$.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the subgiant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

We conducted orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}, but without a joint Gaussian Process modeling procedure \citep{2023ApJ...950..162T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\ms$, a few times larger than the typical Keck HIRES measurements, due to the lower information content in the near-IR than the visible for this relatively rapidly rotating F star.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued.  The period and $T_0$ were fixed from the \emph{TESS} transits, and we assume a circular orbit with $K>10$ \ms.  The semiamplitude prior acts as a mass constraint, excluding planets with masses so low that the observed radius would exceed the Hill radius (Roche lobe overflow).  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \ms, a relatively weak constraint, but consistent with broadly Saturn-mass planets.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  The semiamplitude prior accounts for the apparent improvement in the mass constraint.  Hypothetically, a truly Roche lobe overflow planet could be consistent with the available data, so our mass constraint could be considered an upper limit.

\subsection{HPF analysis II: \ion{He}{1} 10833 \AA} \label{secHeAnalysis}
The preparation for analyzing the Helium feature followed a similar but slightly different procedure than that used for the RV, namely, we used \texttt{Goldilocks}\footnote{\url{https://github.com/grzeimann/Goldilocks_Documentation}} for 2D \'echellogram reduction.  This tool outputs 1D extracted spectra for the target, and two reference fibers: blank sky and a laser frequency comb \citep[LFC,][]{2019Optic...6..233M}.  The observations were acquired with the LFC turned off, so this unilluminated spectrum was discarded.

@@

The HET's fixed-altitude design means that the airmass remains relatively constant across all pointings, lowering the telluric line variability compared to fully steerable telescopes which may sample a wider range in airmasses.  Variability in atmospheric conditions still makes it difficult to mitigate telluric lines to within the photon noise limit, so we masked spectral regions predicted to have significant telluric lines with a template generated by \texttt{telfit} \citep{2014AJ....148...53G}.  We then shifted the spectral coordinates to their common barycentric-corrected reference frame \citep{2014PASP..126..838W}, as implemented in \texttt{astropy} \citep{2013A&A...558A..33A,2018AJ....156..123A,2022ApJ...935..167A}.  The spectral continuum was flattened and normalized to two pre-selected continuum indices highlighted as vertical blue bands in Figure \ref{fig:HPFheliumOverview}.  The sky-subtraction, telluric masking, barycentric correction, flattening, and all of our other standard pre-processing steps were carried out in the open-source Python interface \texttt{muler} \citep{2022JOSS....7.4302G}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/HAT_P_67b_He_spectrum.png}
    \caption{Overlay of all 152 individual HPF exposures of HAT-P-67, spanning 2020-2022. Variability is seen in the \ion{He}{1} 10833 \AA~ triplet near the vertical orange shaded band, but not in the adjacent Si line.  These snapshot spectra were barycentric corrected, and continuum flattened with a linear fit to the regions in the blue vertical bands.  Sharp telluric absorption lines have been masked in regions near 10835 and 10837.5 \AA.}
    \label{fig:HPFheliumOverview}
\end{figure}

Figure \ref{fig:HPFperCampaign} shows the spectra for four campaigns, with before and after spectra showing conspicuous excess absorption during transit and 1 day before transit, but negligible excess absorption after transit.  The individual transits show significant morphological variation, with substructure in the bulk line-of-sight velocity distribution.

@@

The in-transit phases exhibit over 10\% excess absorption peaks.  A large absorption signal can be seen preceding transit ingress, with some significant absorption evident before $-0.2$ in phase.   The egress drop-off is extremely sharp, with almost no excess absorption directly after transit, as seen in the Equivalent Width (EW) timeseries (Figure \ref{fig:HPFtimeseries}).  We construct a fractional residual absorption spectrum by subtracting off and re-normalizing to the non-varying baseline.  We define the non-varying baseline spectrum as the average over phases $0.4-0.5$, which exhibited stable spectra with the least absorption.  Figure \ref{fig:HPFscanResid} shows that significant absorption can be seen as a few percent at $-0.37$ in phase, rising to 10\% just before and during transit. The abrupt dropoff in helium excess at planet egress is conspicuous.

@@

The underlying structure of the metastable \ion{He}{1} triplet consists of three quantum components, two of which ($J=$ 1 and 2) are typically blended owing to Doppler broadening from the finite temperature of the gas $T_0$.  HAT-P-67~b exhibits blending of all three components ($J=$ 0, 1, and 2) into a single broad Gaussian-like feature seen in Figure \ref{fig:HPFheliumOverview}.  The mere observation of this broadening implies that HAT-P-67~b probes a larger velocity dispersion than typical measurements, and may be associated with either a higher gas temperature, complex planetary wind flows, or some mix of both.  Whatever the cause, we treat the feature as a single Gaussian line for the pupose of estimating bulk characteristics of the Helium excess absorption feature.  We restricted the fitting to the 10828$-$10838 \AA~region of the residual spectra.  The model constructed in this way has a total of 4 parameters: amplitude $A$, location $\lambda_c$, Gaussian width $\sigma$, and constant offset.  We repeated a similar process for the nearby Si line at 10830 \AA~as a control sample.

@@

At the moment of transit midcenter, the centroids exhibit a redshifted absorption relative to the planet restframe. Measurements from April 2022 and May 2020 transits partially overlap in phase coverage while yielding slightly different line centroid locations.  Both time series sequences appear consistent with near-zero bulk restframe velocity, but with significant variability in the line profile substructure.  The May 2020 sequence shows a slight trend towards alignment with the planet restframe.  The April 2020 and June 2020 partial transits sampled phases close to planet egress, with a clear blueshift relative to both the star and planet.

Overall this pattern of velocity centroids appears consistent with the majority of the gas rapidly obeying the stellar restframe, but with a large spread in the line-of-sight velocity distribution as we would expect from a range of launch angles.  We revisit the interpretations and caveats for this substructure in the next sections.

@@

We conclude that reinflation likely does not have a significant effect on the evolution of HAT-P-67~b: inflation timescales remain relatively unchanged over the planet lifetime, despite a 50\% spike in incident radiation in the subgiant scenario.  The reason is subtle, as we describe next.

>>>>

\begin{abstract}
    Atmospheric escape shapes the fate of exoplanets, with statistical evidence for transformative mass loss imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10833 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of up to 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67~b. The 13.8 hours of on-sky integration time over 39 nights sample the entire planet orbit, uncovering excess Helium absorption preceding the transit by up to 130 planetary radii in a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary winds.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We contrast two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

@@

Atmospheric escape appears to be a normal phase for exoplanets of a certain size and insolation, statistically imprinted in the dearth of 1.5$-$2.0~$R_\oplus$ planets within 100-day periods \citep{2017AJ....154..109F}.  The leading candidate physical mechanisms for this rapid transition between mini-Neptunes and super-Earths include photoevaporation \citep{2013ApJ...775..105O,2013ApJ...776....2L,2017ApJ...847...29O} and core-powered mass loss \citep{2019MNRAS.487...24G,2023arXiv230200009B}.  Whatever the cause, some large fraction of planets undergo transformative atmospheric escape, and the signal should be widely discernable.  Such signals have been searched for, and increasingly found, in many transiting planet systems with at least 28 detections to date \citep{2022arXiv221116243D}.

@@

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may either form \emph{in-situ} or effectively migrate to close-in orbital separations \citep{2018ARA&A..56..175D}, but once they arrive, the intense irradiation overheats the planet.  This heating leads to runaway inflation and, ultimately, complete disintegration.  In this scenario, the inflationary half-life becomes so short that the probability of observing members in the class decreases sharply with density, causing the apparent lack of inflated sub-Saturns \citep{2023ApJ...945L..36T}.

@@

Recently, \citet{2023ApJ...945L..36T} showed that hot Saturns can undergo catastrophic erosion by stellar X-ray and Extreme UV (XUV) photoevaporative mass loss. Planets with densities less than $\sim$0.1 g~cm$^{-3}$ achieve mass loss rates of up to $10^3$ $M_\oplus /$Gyr, setting up a positive feedback loop: the planet increases in radius, overflowing its Roche lobe, fueling greater mass loss, and increasing in radius even further.  This vicious cycle systematically depopulates the mass-radius plane with a cliff defined by the $\rho_p \sim$0.1 g~cm$^{-3}$ dividing line.

@@

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2000ApJ...537..916S,2003Natur.422..143V,2012ApJ...751...86J,2018NatAs...2..714Y,2018ApJ...855L..11O,2018Natur.557...68S, 2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections \citep{2022arXiv221116243D}\footnote{See also Table~S1 of \citealt{doi:10.1126/sciadv.adf8736} for a compilation of exoplanets with detections and non-detections of the helium excess.}.  The sample of detections includes both hot Jupiters and lower mass planets but lacks inflated hot Saturns owing to their intrinsic rarity below the 0.1 g~cm$^{-3}$ threshold. An observational picture of mass loss in inflated hot Saturns appears to be lacking for this reason.

@@

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis}, we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in Section \ref{secHeAnalysis}, with an analysis of the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere with 1D Parker wind models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradiation and Ohmic dissipation.  Finally, in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

The Habitable Zone Planet Finder Spectrograph \citep[HPF;][]{2012SPIE.8446E..1SM,2014SPIE.9147E..1GM, 2019Optic...6..233M} on the queue-scheduled 10-meter \emph{Hobby-Eberly Telescope} \citep[HET;][]{1998SPIE.3352...34R} operates in the near-IR from $8100-12800~$\AA~ spanning the \textit{z}, \textit{Y}, and \textit{J} bands at spectral resolving power $R=55,000$. The HET fixed-elevation design \citep{2007PASP..119..556S} limits the observability of \object{HAT-P-67} to less than 1 hour ``tracks'' for a fixed range of hour angles before (east track) and after (west track) the star transits the meridian.  Whereas conventional steerable telescopes could conduct continuous point-and-stare observations of \object{HAT-P-67} for hours, HET cannot.  In practice, this limitation means that in-transit and out-of-transit observational phases were rarely possible on the same night.  Instead, we organized the observations into four campaigns to coincide with \object{HAT-P-67 b} transits on UT dates 2020 April 28, 2020; May 22, 2020; June 15, 2020; and 2022 April 29.  These campaigns have out-of-transit observations at least one night before and one night after, and often two nights on either side of the transit.  Two more transit snapshots were obtained in 2022 June-July without the accompanying visits immediately before and after.  The in-transit campaigns had up to 14 exposures per HET track, with integration times between 5-8.5 minutes.  We also obtained random-in-phase lower-priority ``P1-P4'' reconnaissance observations \citep{2007PASP..119..556S}.  These out-of-transit snapshot observations typically received 4 or fewer individual exposures.  We observed HAT-P-67 with HPF for a total of 41 visits on 39 unique nights, with two of those nights observing both the east and west HET tracks.  The total on-source integration time exceeds 13.8 hours, with $N=152$ individual spectra possessing a typical signal-to-noise ratio of 80 per pixel.  Table \ref{tabHPFLog} summarizes the log of observations.

The observation schedule was strategically restricted to the spring season when the Barycentric Earth Radial Velocity (BERV) would Doppler shift telluric absorption lines sufficiently far away from the core of the Helium 10833 \AA~ feature \citep{2022ApJ...939L..11S}.  The small BERV still means the redward Helium line wing has significant telluric contamination between 10834$-$10836 \AA, but the core and blue line wings appear relatively pristine.  Sky emission lines land in this region but are more easily mitigated by HPF's simultaneous sky reference fiber.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}, well-separated from the planet-host star and not a source of contamination for the HPF observations.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ mag source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity since the wide companion will not factor into our analysis.

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation. TESS Sectors 51-53 do not show as conspicuous a modulation signal but still show some ostensibly stellar variability.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone, the periods hover around 5.9 days.  The signal is weak enough, especially in Sectors 51-53, that \emph{TESS} instrumental systematics may contribute some of the out-of-transit modulation that we see in Figure \ref{fig:TESSoverview}.

We can independently constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i_\star}$, and the observation of the planet's orbital inclination $i_p\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We assume spin-orbit alignment, $i_\star \sim i_p$.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the subgiant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

We conducted orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}, but without a joint Gaussian Process modeling procedure \citep{2023ApJ...950..162T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\ms$, a few times larger than the typical Keck HIRES measurements, due to the lower information content in the near-IR than the visible for this relatively rapidly rotating F star.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued.  The period and $T_0$ were fixed from the \emph{TESS} transits, and we assume a circular orbit with $K>10$ \ms.  The semiamplitude prior acts as a mass constraint, excluding planets with masses so low that the observed radius would exceed the Hill radius (Roche lobe overflow).  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \ms, a relatively weak constraint but consistent with broadly Saturn-mass planets.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  The semiamplitude prior accounts for the apparent improvement in the mass constraint.  Hypothetically, a truly Roche lobe overflow planet could be consistent with the available data, so our mass constraint could be considered an upper limit.

\subsection{HPF analysis II: \ion{He}{1} 10833 \AA} \label{secHeAnalysis}
The preparation for analyzing the Helium feature followed a similar but slightly different procedure than that used for the RV. Namely, we used \texttt{Goldilocks}\footnote{\url{https://github.com/grzeimann/Goldilocks_Documentation}} for 2D \'echellogram reduction.  This tool outputs 1D extracted spectra for the target and two reference fibers: blank sky and a laser frequency comb \citep[LFC,][]{2019Optic...6..233M}.  The observations were acquired with the LFC turned off, so this unilluminated spectrum was discarded.

@@

The HET's fixed-altitude design means that the airmass remains relatively constant across all pointings, lowering the telluric line variability compared to fully steerable telescopes, which may sample a wider range of airmasses.  Variability in atmospheric conditions still makes it difficult to mitigate telluric lines to within the photon noise limit, so we masked spectral regions predicted to have significant telluric lines with a template generated by \texttt{telfit} \citep{2014AJ....148...53G}.  We then shifted the spectral coordinates to their common barycentric-corrected reference frame \citep{2014PASP..126..838W}, as implemented in \texttt{astropy} \citep{2013A&A...558A..33A,2018AJ....156..123A,2022ApJ...935..167A}.  The spectral continuum was flattened and normalized to two pre-selected continuum indices highlighted as vertical blue bands in Figure \ref{fig:HPFheliumOverview}.  The sky-subtraction, telluric masking, barycentric correction, flattening, and all of our other standard pre-processing steps were carried out in the open-source Python interface \texttt{muler} \citep{2022JOSS....7.4302G}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/HAT_P_67b_He_spectrum.png}
    \caption{Overlay of all 152 individual HPF exposures of HAT-P-67, spanning 2020-2022. Variability is seen in the \ion{He}{1} 10833 \AA~ triplet near the vertical orange shaded band but not in the adjacent Si line.  These snapshot spectra were barycentric corrected and continuum flattened with a linear fit to the regions in the blue vertical bands.  Sharp telluric absorption lines have been masked in regions near 10835 and 10837.5 \AA.}
    \label{fig:HPFheliumOverview}
\end{figure}

Figure \ref{fig:HPFperCampaign} shows the spectra for four campaigns, with before and after spectra showing conspicuous excess absorption during transit and 1 day before transit but negligible excess absorption after transit.  The individual transits show significant morphological variation, with substructure in the bulk line-of-sight velocity distribution.

@@

The in-transit phases exhibit over 10\% excess absorption peaks.  A large absorption signal can be seen preceding transit ingress, with some significant absorption evident before $-0.2$ in phase.   The egress drop-off is extremely sharp, with almost no excess absorption directly after transit, as seen in the Equivalent Width (EW) time series (Figure \ref{fig:HPFtimeseries}).  We construct a fractional residual absorption spectrum by subtracting off and re-normalizing to the non-varying baseline.  We define the non-varying baseline spectrum as the average over phases $0.4-0.5$, which exhibited stable spectra with the least absorption.  Figure \ref{fig:HPFscanResid} shows that significant absorption can be seen as a few percent at $-0.37$ in phase, rising to 10\% just before and during transit. The abrupt dropoff in helium excess at planet egress is conspicuous.

@@

The underlying structure of the metastable \ion{He}{1} triplet consists of three quantum components, two of which ($J=$ 1 and 2) are typically blended owing to Doppler broadening from the finite temperature of the gas $T_0$.  HAT-P-67~b exhibits blending of all three components ($J=$ 0, 1, and 2) into a single broad Gaussian-like feature seen in Figure \ref{fig:HPFheliumOverview}.  The mere observation of this broadening implies that HAT-P-67~b probes a larger velocity dispersion than typical measurements and may be associated with either a higher gas temperature, complex planetary wind flows, or some mix of both.  Whatever the cause, we treat the feature as a single Gaussian line for the purpose of estimating bulk characteristics of the Helium excess absorption feature.  We restricted the fitting to the 10828$-$10838 \AA~region of the residual spectra.  The model constructed in this way has a total of 4 parameters: amplitude $A$, location $\lambda_c$, Gaussian width $\sigma$, and constant offset.  We repeated a similar process for the nearby Si line at 10830 \AA~as a control sample.

@@

At the moment of transit midcenter, the centroids exhibit a redshifted absorption relative to the planet restframe. Measurements from April 2022 and May 2020 transits partially overlap in phase coverage while yielding slightly different line centroid locations.  Both time series sequences appear consistent with near-zero bulk restframe velocity but with significant variability in the line profile substructure.  The May 2020 sequence shows a slight trend towards alignment with the planet restframe.  The April 2020 and June 2020 partial transits sampled phases close to planet egress, with a clear blueshift relative to both the star and planet.

Overall this pattern of velocity centroids appears consistent with the majority of the gas rapidly obeying the stellar restframe but with a large spread in the line-of-sight velocity distribution, as we would expect from a range of launch angles.  We revisit the interpretations and caveats for this substructure in the next sections.

@@

We conclude that reinflation likely does not have a significant effect on the evolution of HAT-P-67~b: inflation timescales remain relatively unchanged over the planet's lifetime, despite a 50\% spike in incident radiation in the subgiant scenario.  The reason is subtle, as we describe next.
